Write-and-f-array: implementation and an application

### 2013-09-12 msc.tex
formatted, pgbroken

\documentclass[a4paper,11pt]{article}
%\usepackage{algorithmicx}
\usepackage{fancyvrb}
\usepackage{fullpage}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphics}
\usepackage{varwidth}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{placeins}
\usepackage{amsthm}
\usepackage[noend]{algpseudocode}
\usepackage{tikz}
\usepackage{comment}
\usetikzlibrary{arrows,decorations.pathreplacing}
\newcommand{\atomic}[4]{\node(lab#1) at (#3,-#4+0.5) {#2};\draw (#3,-#4-0.1) to (lab#1);\node[inner sep=0pt](#1) at (#3,-#4) {};}
\algblockdefx{Struct}{EndStruct}[1]{\textbf{struct} \textsc{#1}}{}
\algnewcommand{\algorithmicgoto}{\textbf{go to}}
\algnewcommand{\Goto}{\textbf{goto} }
\algnewcommand{\Label}{\State\unskip}
\def\none{\textbf{none} }
\def\T{\ensuremath{\operatorname{\mathcal{T}}}\text{ }}
\def\int{\ensuremath{\operatorname{\textbf{int}}}}
\algnotext{EndStruct}
\newtheorem{lemma}{Lemma}
\newtheorem{observation}{Observation}
\newtheorem{theorem}{Theorem}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}
\newcommand{\code}[1]{\text{#1}}
\newcommand{\fn}[1]{\textsc{#1}}
\newcommand{\var}[2]{\textbf{var }#1 : #2}
\newcommand{\arrayspec}[1]{\text{array}[#1]\text{ of }}
\usepackage{float}
\newfloat{Listing}{t}{lop}
\author{Robert Obryk}
\title{Write-and-f-array: implementation and an application}
\date{}
\VerbatimFootnotes
\begin{document}

@@

The write-and-f-array can be applied to create a fetch-and-add object for $P$ processes with $O(P \log P)$ memory complexity and $O(\log^3 P)$ step complexity of the fetch-and-add operation. This is the first implementation of fetch-and-add with polylogarithmic step complexity and subquadratic memory complexity that can be implemented without CAS or LL/SC objects of unrealistic size\cite{ellen-fai}.
\end{abstract}
\clearpage
\tableofcontents
\clearpage
\section{Introduction}

@@

\subsection{Concurrent Objects}
A useful abstraction in a single-threaded system is an object: an entity with a set of methods one can invoke,
and a specification of semantics of these methods. Naturally, a program can use multiple objects and invoke
their methods in any order.

@@

\Function{wr}{$y$}
	\State $x \gets y$
\EndFunction

\Function{CAS}{$o, n$}
\If{$x = o$}
	\State $x \gets n$
	\State\Return true
\Else
	\State\Return false
\EndIf
\EndFunction
\end{algorithmic}

@@

\subsection{Serialization}
In many situations it would be convenient to have more complicated concurrent objects: counters that can be incremented,
queues that can be pushed to and popped from, etc. Alas, real-world hardware usually doesn't implement such objects
directly. We will thus implement them using CAS registers or other simpler objects: we will substitute a set of simpler objects for every such object and for every method call on this object, execute a procedure that implements it instead.
However, these prodecures might run concurrently, so we can't blindly use a single-threaded implementation of the object.
Obviously, we will want such substitution to preserve semantics of the program in which it was substituted. The following condition immediately implies
this\footnote{For a detailed discussion of subtleties related to this definition, refer to chapter 3.3-3.6 of \cite{TheArt}}:

@@

\subsection{Wait-freedom}

@@

\subsection{Snapshots}

@@

\section{The History Object}

@@

\begin{algorithmic}[1]
	\State\var{$S$}{$\textbf{pair}\left<v: \int, p: \int\right>$} \Comment{Version number and $p$ of the most recently published value}
	\State\var{$H$}{\arrayspec{N}$\textbf{pair}\left<v: \int, T: \T\right>$} \Comment{A circular buffer of recent $N$ values}
	\State\var{$L$}{\arrayspec{P}$\textbf{pair}\left<v: \int, T: \T\right>$} \Comment{Temporary storage for values being published}
	\Function{get-current}{}
		\State $s \gets S.\fn{rd}()$ \Comment{Possible serialization point}
		\State \fn{help}()
		\State \Return $H[s.v \mod N].\fn{rd}()$
	\EndFunction
	\Function{help}{} \Comment{Updates $H[]$ with value from $L[]$, if required}
		\State $s \gets S.\fn{rd}()$
		\State $l \gets L[s.p].\fn{rd}()$
		\State $h \gets H[s.v \mod N].\fn{rd}()$
		\If{$l.v = s.v \textbf{ and } h.v < s.v$}
			\State $H[s.v \mod N].\fn{CAS}(h, l)$ \label{hist-help-cas}
		\EndIf
	\EndFunction
	\Function{get}{v}
		\State $s \gets S.\fn{rd}()$ \label{hist-get-notyet-sp} \Comment{Possible serialization point}
		\If{$s.v < v$}
			\State \Return \none \label{hist-get-notyet} \Comment{The requested version hasn't been published yet}
		\EndIf
		\State \fn{help}()
		\State $h \gets H[v \mod N].\fn{rd}()$
		\If{$h.v = v$}
			\State \Return $h.T$
		\Else
			\State \Return \none \label{hist-get-tooold}
		\EndIf
	\EndFunction
	\Function{publish$_p$}{$v, T$}
		\State $s \gets S.\fn{rd}()$ \label{hist-pub-earlier-sp} \Comment{Possible serialization point}
		\If{$v \neq s.v + 1$}
			\State \Return \textbf{false} \label{hist-pub-earlier-exit}
		\EndIf
		\State \fn{help}()
		\State $L[p].\fn{wr}((v, T))$ \label{hist-pub-latest}
		\If{$S.\fn{CAS}(s, (v, p)$) \textbf{failed}} \Comment{Possible serialization point}\label{hist-pub-later-sp}
			\State \Return \textbf{false}
		\EndIf
		\State \Return \textbf{true}
	\EndFunction
\end{algorithmic}

@@

Obviously all operations run in $O(1)$ time. Memory complexity is $O(N+P)$, where $N$ and $P$ are the parameters defined at the beginning of this section.
We will now use this object in the main result of this work, an implementation of a write-and-$f$-array.
\section{The Write-and-f-array}

@@

\section{Implementation}
A fetch-and-add object was implemented using the above-mentioned construction from the write-and-f object. The implementation is written in C++ and uses the C++11 standard library support for atomic operations. It is published on Github \footnote{The version current as of the time this work is written is available at: \verb+https://github.com/robryk/parsum/tree/d3433f7f7b137b52a73cfb244d46081528c696c3+ }.

@@

\section{Extensions and Open Problems}

@@

We pose two problems:

@@

\begin{comment}
\section{Conclusions}
The fetch-and-f-array was introduced and its wait-free implementation with polylogarithmic step complexities
and near-linear memory complexity was provided. This in turn was used to produce a fetch-and-add counter
with step complexity polylgarithmic in number of threads and memory complexity nearly linear in number
of threads. The counter was implemented on real hardware, benchmarked, and found to be between 10 and 100
times slower than a trivial lock-free counter. A possible constant-factor improvement was outlined, but
not investigated. Some open problems were stated.
\end{comment}

\section{Acknowledgements}
The author would like to thank his advisor, dr. Grzegorz Herman for many fruitful discussions and help while preparing this paper. He would also like to thank Szymon Acedański for his help in improving the presentation of this paper.

>>>>

%\documentclass[a4paper,11pt]{article}
\documentclass{thesis}
%\usepackage{algorithmicx}
\usepackage{fancyvrb}
\usepackage{fullpage}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphics}
\usepackage{varwidth}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{placeins}
\usepackage{amsthm}
\usepackage[noend]{algpseudocode}
\usepackage{tikz}
\usepackage{comment}
\usetikzlibrary{arrows,decorations.pathreplacing}
\newcommand{\atomic}[4]{\node(lab#1) at (#3,-#4+0.5) {#2};\draw (#3,-#4-0.1) to (lab#1);\node[inner sep=0pt](#1) at (#3,-#4) {};}
\algblockdefx{Struct}{EndStruct}[1]{\textbf{struct} \textsc{#1}}{}
\algnewcommand{\algorithmicgoto}{\textbf{go to}}
\algnewcommand{\Goto}{\textbf{goto} }
\algnewcommand{\Label}{\State\unskip}
\def\none{\textbf{none} }
\def\T{\ensuremath{\operatorname{\mathcal{T}}}\text{ }}
\def\int{\ensuremath{\operatorname{\textbf{int}}}}
\algnotext{EndStruct}
\newtheorem{lemma}{Lemma}
\newtheorem{observation}{Observation}
\newtheorem{theorem}{Theorem}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}
\newcommand{\code}[1]{\text{#1}}
\newcommand{\fn}[1]{\textsc{#1}}
\newcommand{\var}[2]{\textbf{var }#1 : #2}
\newcommand{\arrayspec}[1]{\text{array}[#1]\text{ of }}
\usepackage{float}
\newfloat{Listing}{t}{lop}
\author{Robert Obryk}
\title{Write-and-f-array: implementation and an application}
\nralbumu{1040504}
\kierunek{Computer Science - IT Analyst}
\opiekun{dr Grzegorz Herman}
\date{September 2013}
\VerbatimFootnotes
\begin{document}

@@

The write-and-f-array can be applied to create a fetch-and-add object for $P$ processes with $O(P \log P)$ memory complexity and $O(\log^3 P)$ step complexity of the fetch-and-add operation. This is the first implementation of fetch-and-add with polylogarithmic step complexity and subquadratic memory complexity that can be implemented without CAS or LL/SC objects of unrealistic size\cite{ellen-fai}.
\end{abstract}
\clearpage
\tableofcontents
\clearpage
\chapter{Introduction}

@@

\section{Concurrent Objects}
A useful abstraction in a single-threaded system is an object: an entity with a set of methods one can invoke,
and a specification of semantics of these methods. Naturally, a program can use multiple objects and invoke
their methods in any order.

@@

\Function{wr}{$y$}
	\State $x \gets y$
\EndFunction
\pagebreak
\Function{CAS}{$o, n$}
\If{$x = o$}
	\State $x \gets n$
	\State\Return true
\Else
	\State\Return false
\EndIf
\EndFunction
\end{algorithmic}

@@

\section{Serialization}
In many situations it would be convenient to have more complicated concurrent objects: counters that can be incremented,
queues that can be pushed to and popped from, etc. Alas, real-world hardware usually doesn't implement such objects
directly. We will thus implement them using CAS registers or other simpler objects: we will substitute a set of simpler objects for every such object and for every method call on this object, execute a procedure that implements it instead.
However, these prodecures might run concurrently, so we can't blindly use a single-threaded implementation of the object.
Obviously, we will want such substitution to preserve semantics of the program in which it was substituted. The following condition immediately implies
this\footnote{For a detailed discussion of subtleties related to this definition, refer to chapter 3.3-3.6 of \cite{TheArt}}:

@@

\section{Wait-freedom}

@@

\section{Snapshots}

@@

\chapter{The History Object}

@@

\begin{algorithmic}[1]
	\State\var{$S$}{$\textbf{pair}\left<v: \int, p: \int\right>$} \Comment{Version number and $p$ of the most recently published value}
	\State\var{$H$}{\arrayspec{N}$\textbf{pair}\left<v: \int, T: \T\right>$} \Comment{A circular buffer of recent $N$ values}
	\State\var{$L$}{\arrayspec{P}$\textbf{pair}\left<v: \int, T: \T\right>$} \Comment{Temporary storage for values being published}
	\Function{get-current}{}
		\State $s \gets S.\fn{rd}()$ \Comment{Possible serialization point}
		\State \fn{help}()
		\State \Return $H[s.v \mod N].\fn{rd}()$
	\EndFunction
	\Function{help}{} \Comment{Updates $H[]$ with value from $L[]$, if required}
		\State $s \gets S.\fn{rd}()$
		\State $l \gets L[s.p].\fn{rd}()$
		\State $h \gets H[s.v \mod N].\fn{rd}()$
		\If{$l.v = s.v \textbf{ and } h.v < s.v$}
			\State $H[s.v \mod N].\fn{CAS}(h, l)$ \label{hist-help-cas}
		\EndIf
	\EndFunction
	\pagebreak
	\Function{get}{v}
		\State $s \gets S.\fn{rd}()$ \label{hist-get-notyet-sp} \Comment{Possible serialization point}
		\If{$s.v < v$}
			\State \Return \none \label{hist-get-notyet} \Comment{The requested version hasn't been published yet}
		\EndIf
		\State \fn{help}()
		\State $h \gets H[v \mod N].\fn{rd}()$
		\If{$h.v = v$}
			\State \Return $h.T$
		\Else
			\State \Return \none \label{hist-get-tooold}
		\EndIf
	\EndFunction
	\Function{publish$_p$}{$v, T$}
		\State $s \gets S.\fn{rd}()$ \label{hist-pub-earlier-sp} \Comment{Possible serialization point}
		\If{$v \neq s.v + 1$}
			\State \Return \textbf{false} \label{hist-pub-earlier-exit}
		\EndIf
		\State \fn{help}()
		\State $L[p].\fn{wr}((v, T))$ \label{hist-pub-latest}
		\If{$S.\fn{CAS}(s, (v, p)$) \textbf{failed}} \Comment{Possible serialization point}\label{hist-pub-later-sp}
			\State \Return \textbf{false}
		\EndIf
		\State \Return \textbf{true}
	\EndFunction
\end{algorithmic}

@@

Obviously all operations run in $O(1)$ time. Memory complexity is $O(N+P)$, where $N$ and $P$ are the parameters defined at the beginning of this section.
We will now use this object in the main result of this work, an implementation of a write-and-$f$-array.
\chapter{The Write-and-f-array}

@@

\chapter{Implementation}
A fetch-and-add object was implemented using the above-mentioned construction from the write-and-f object. The implementation is written in C++ and uses the C++11 standard library support for atomic operations. It is published on Github \footnote{The version current as of the time this work is written is available at: \verb+https://github.com/robryk/parsum/tree/d3433f7f7b137b52a73cfb244d46081528c696c3+ }.

@@

\chapter{Extensions and Open Problems}

@@

We pose also two purely theoretical problems:

@@

\begin{comment}
\chapter{Conclusions}
The fetch-and-f-array was introduced and its wait-free implementation with polylogarithmic step complexities
and near-linear memory complexity was provided. This in turn was used to produce a fetch-and-add counter
with step complexity polylgarithmic in number of threads and memory complexity nearly linear in number
of threads. The counter was implemented on real hardware, benchmarked, and found to be between 10 and 100
times slower than a trivial lock-free counter. A possible constant-factor improvement was outlined, but
not investigated. Some open problems were stated.
\end{comment}

\begingroup\let\clearpage\relax\chapter{Acknowledgements}\endgroup
The author would like to thank his advisor, dr. Grzegorz Herman for many fruitful discussions and help while preparing this paper. He would also like to thank Szymon Acedański for his help in improving the presentation of this paper.
